---
title: "Gut GVHD Pilot"
format:
  html:
    embed-resources: true
editor: visual
execute:
  echo: true
  output: true
  warning: false
  message: false
---

#### Data Loading

Sample SLV14 was chosen because it is an example of severe GVHD. 16 micrometer resolution was chosen due to limitations of computational resources.

```{r}
library(Seurat)
library(dplyr)

matrix_dir <- "data/SLV14/binned_outputs/square_002um/filtered_feature_bc_matrix"
spatial_dir <- "data/SLV14/binned_outputs/square_002um/spatial"

counts <- Read10X(data.dir = matrix_dir)
slv14 <- CreateSeuratObject(counts, assay = "Spatial", project = "SLV14")
slv14 <- subset(slv14, subset = nCount_Spatial >= 50)

img <- Read10X_Image(image.dir = spatial_dir, image.name = "tissue_lowres_image.png")
img <- img[Cells(slv14)]
DefaultAssay(img) <- "Spatial"
slv14[["slice1"]] <- img  # attach image to the object

slv14
```

#### Data Quality Control

I used Median Absolute Deviations as my standard for filtering spots. Data poitns above/below 3 MADs for any of the criterion (features, counts, percent.mt) were excluded from the dataset.

```{r}

slv14[["percent.mt"]] <- PercentageFeatureSet(slv14, pattern = "^MT-")

VlnPlot(slv14, features = c("nCount_Spatial", "nFeature_Spatial", "percent.mt"))
FeatureScatter(slv14, feature1 = "nCount_Spatial", feature2 = "nFeature_Spatial")

log_count <- log1p(slv14$nCount_Spatial)
med_count <- median(log_count)
mad_val_count <- mad(log_count)

count_lower_log <- (med_count - (mad_val_count*3))
count_upper_log <- (med_count + (mad_val_count*3))

count_lower <- expm1(count_lower_log)
count_upper <- expm1(count_upper_log)

log_feature <- log1p(slv14$nFeature_Spatial)
med_feature <- median(log_feature)
mad_val_feature <- mad(log_feature)

feature_lower_log <- (med_feature - (mad_val_feature*3))
feature_upper_log <- (med_feature + (mad_val_feature*3))

feature_lower <- expm1(feature_lower_log)
feature_upper <- expm1(feature_upper_log)

mt_upper = median(slv14$percent.mt) + (3 * mad(slv14$percent.mt))

dim(slv14)

slv14 <- subset(slv14, subset = nCount_Spatial < count_upper & nCount_Spatial > count_lower & nFeature_Spatial > feature_lower & nFeature_Spatial < feature_upper & percent.mt < mt_upper)

VlnPlot(slv14, features = c("nCount_Spatial", "nFeature_Spatial", "percent.mt"))
FeatureScatter(slv14, feature1 = "nCount_Spatial", feature2 = "nFeature_Spatial")

dim(slv14)
```

#### Data Normalization and Scaling

SCTransform() was chosen to normalize and scale the dataset.

```{r}
slv14 <- SCTransform(slv14, assay = "Spatial", vars.to.regress = "percent.mt",
                     do.correct.umi = FALSE, conserve.memory = TRUE)
```

#### Data Dimension Reduction

#TODO: A lower clustering resolution was used after FindAllMarkers() returned no features for two clusters. Even after lower resolution was used, one cluster still returned empty. Revisit.

```{r}
slv14 = RunPCA(slv14, assay = "SCT", npcs = 50)

ElbowPlot(slv14, ndims = 50)

slv14 <- FindNeighbors(slv14, reduction = "pca", dims = 1:15)
slv14 <- FindClusters(slv14, resolution = 0.4)

slv14 <- RunUMAP(slv14, dims = 1:15)

DimPlot(slv14, reduction = "umap", label = TRUE)
SpatialDimPlot(slv14, label = TRUE, label.size = 3)


```

#### Data Markers

#TODO: Revisit empty cluster 0.

```{r}
slv14_markers <- FindAllMarkers(slv14, only.pos = TRUE, min.pct = 0.10, logfc.threshold = 0.25)

slv14_top_markers <- slv14_markers %>% filter(p_val_adj < 0.05, pct.1 - pct.2 > 0.10) %>% group_by (cluster) %>% slice_max(avg_log2FC, n = 10)

print(slv14_top_markers)
```

#### Data Module Scoring

Module scoring was used to determine transcription activity for each bin population profile quantitatively.

```{r}
cd8_eff_genes <- c("CD3E","CD3G","CD3D","NKG7","CST7","GZMK","GZMH","CD8A","GZMB","GZMM","GZMA","PRF1","LYST","LAG3","APOBEC3G")
cd4_eff_genes <- c("CD3E","CD3G","CD3D","CD40LG","CD4","TNFRSF4","IL7R","CCR4","CD28","TNF","IL2","DYNLT3","DPP4","IL4I1","CCL20")
isc_genes     <- c("LGR5","ASCL2","SMOC2","MSI1","RARRES2","SOX9","CENPM","EPHB2","OLFM4")

slv14 <- AddModuleScore(slv14, features = list(cd8_eff_genes), name = "CD8eff_score")
slv14 <- AddModuleScore(slv14, features = list(cd4_eff_genes), name = "CD4eff_score")
slv14 <- AddModuleScore(slv14, features = list(isc_genes),     name = "ISC_score")

SpatialFeaturePlot(slv14, features = c("CD8eff_score1", "CD4eff_score1", "ISC_score1"))

```
